---
title: "Sedgwick & SBFR netCDF pre-processing and spatial base station creation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load libraries
```{r, warning=FALSE, results=FALSE}
library(ncdf4)
library(raster)
library(sp)
library(sf)
library(maps)
library(terra)
library(sp)
#library(ncdf4.helpers)
library(ggplot2)
library(tidyverse)
library(FNN) 
```

# Sedgwick

## build a lon/lat raster to convert WRF coordinates
The fix from previous attempts: build a proper regular lon/lat raster over the basin, and for every cell of that regular raster look up the nearest real WRF grid cell (by true lon/lat distance, not by array index). Do that lookup once, then reuse it to pull every day/every year. This produces a normal, valid, regularly gridded netCDF that `terra`, the C program, and RHESSys itself can all read correctly.

### build a raster
```{r}
# define and reproject Sedgwick basin
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")
basin_vect = as.polygons(basin, dissolve = TRUE, na.rm = TRUE)
basin_ll <- project(basin_vect, "EPSG:4326")

# pull WRF grid geometry directly by dimension name
get_wrf_geom <- function(nc_path) {
  nc <- nc_open(nc_path)
  on.exit(nc_close(nc))
  lon <- ncvar_get(nc, "XLONG")
  lat <- ncvar_get(nc, "XLAT")
  dn  <- sapply(nc$var[["XLONG"]]$dim, function(d) d$name) # c("south_north","west_east")
  n_sn <- nc$dim[["south_north"]]$len
  n_we <- nc$dim[["west_east"]]$len
  
  # ncvar_get preserves the file's declared dimension order, so reshape using dn, not an assumption
  dim(lon) <- if (dn[1] == "south_north") c(n_sn, n_we) else c(n_we, n_sn)
  dim(lat) <- dim(lon)
  if (dn[1] != "south_north") { lon <- t(lon); lat <- t(lat) } # normalize to [south_north, west_east]
  list(lon = lon, lat = lat, n_sn = n_sn, n_we = n_we)
}

# choose example file (any one WRF file works -- grid geometry is assumed static across years/vars)
geom <- get_wrf_geom("/Users/grace/rprojects/wrf_project/dly_prec/wrf.dlytotal.precip.1988.nc")

# flatten with explicit south_north (iy) / west_east (ix) row/col indices
wrf_pts <- data.frame(
  lon = as.vector(geom$lon),
  lat = as.vector(geom$lat),
  iy  = rep(seq_len(geom$n_sn), times = geom$n_we),  # south_north index (row)
  ix  = rep(seq_len(geom$n_we), each  = geom$n_sn)    # west_east index (col)
)
head(wrf_pts)

# convert to terra spatial vector, for a visual basin-overlap sanity check only
wrf_vect <- vect(wrf_pts, geom = c("lon", "lat"), crs = "EPSG:4326")
wrf_sub <- wrf_vect[basin_ll]
plot(wrf_sub, pch = ".", cex = 0.3)
plot(basin_ll, add = TRUE, border = "darkblue", lwd = 2)
```

### build the regridding table
```{r}
# cover the basin's bounding box (in lon/lat) with a small buffer at original WRF resolution
res_deg <- median(c(diff(sort(unique(round(wrf_pts$lat, 5))))[1:20],
                    diff(sort(unique(round(wrf_pts$lon, 5))))[1:20]), na.rm = TRUE)

buf_deg <- res_deg * 2
target_ext <- ext(basin_ll) + buf_deg
target_grid <- rast(target_ext, resolution = res_deg, crs = "EPSG:4326")

# centroid of every target cell
target_pts <- as.points(target_grid, na.rm = FALSE)
target_xy  <- crds(target_pts)

# nearest WRF source point for every target cell (true geographic nearest-neighbor, not index-based)
nn <- FNN::get.knnx(data = as.matrix(wrf_pts[, c("lon","lat")]),
                     query = target_xy, k = 1)
nearest_idx <- nn$nn.index[, 1]   # row into wrf_pts for each target cell, in target-grid cell order

# keep the ix/iy pair each target cell should pull from, reshaped to the target raster's own layout
map_ix <- wrf_pts$ix[nearest_idx]
map_iy <- wrf_pts$iy[nearest_idx]
```

## process the climate data files
The goal of this chunk is to reprocess all of the climate data to ensure that the variables are in the unit and scale expected by the createbasestation.c file and RHESSys itself. 

### create regridding function and function to convert units
```{r}
# build regridding function
regrid_wrf_var <- function(nc_file, varname, target_grid, map_ix, map_iy,
                            is_cumulative = FALSE,
                            convert = c("none", "K_to_C", "mm_to_m")) {
  convert <- match.arg(convert)

  nc <- nc_open(nc_file)
  on.exit(nc_close(nc))
  arr <- ncvar_get(nc, varname)            
  dn  <- sapply(nc$var[[varname]]$dim, function(d) d$name)
  time_pos <- which(dn == "times")
  arr <- aperm(arr, c(which(dn == "south_north"), which(dn == "west_east"), time_pos)) # force [iy, ix, time]

  n_time <- dim(arr)[3]
  n_target <- length(map_ix)

  out_vals <- matrix(NA_real_, nrow = n_target, ncol = n_time)
  for (t in seq_len(n_time)) {
    slice <- arr[, , t]                       # [iy, ix]
    out_vals[, t] <- slice[cbind(map_iy, map_ix)]
  }

  # cumulative -> daily happens before unit conversion. It doesn't matter for
  # precip (mm_to_m is linear, so diff() then divide == divide then diff()),
  # but keeping this order is the safer default if a cumulative variable that
  # needs a non-linear conversion is ever added.
  if (is_cumulative) out_vals <- t(apply(out_vals, 1, function(x) c(x[1], diff(x))))

  out_vals <- switch(convert,
    none    = out_vals,
    K_to_C  = out_vals - 273.15,
    mm_to_m = out_vals / 1000
  )

  # reassemble into the target raster's cell order, one layer per day
  layers <- lapply(seq_len(n_time), function(t) {
    r <- target_grid
    values(r) <- out_vals[, t]
    r
  })
  r_out <- rast(layers)

  times_raw <- ncvar_get(nc, "times")
  time_units <- ncatt_get(nc, "times", "units")$value
  origin <- sub("days since ", "", time_units)
  time(r_out) <- as.Date(origin) + times_raw
  names(r_out) <- format(time(r_out), "wrf_day_%Y%m%d")
  r_out
}
```

### convert and regrid daily cumulative precipitation data
```{r, precip data}
# convert precipitation data
prec_dir <- "/Users/grace/rprojects/wrf_project/dly_prec"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_prec_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
prec_files <- list.files(prec_dir, pattern = "\\.nc$", full.names = TRUE)

all_prec <- lapply(prec_files, regrid_wrf_var, varname = "precsum",
                    target_grid = target_grid, map_ix = map_ix, map_iy = map_iy,
                    is_cumulative = TRUE)

r_basin <- rast(all_prec) # concatenate all years along time
r_basin <- mask(r_basin, basin_ll) # drop cells outside the basin polygon

# sanity check: pick the basin cell nearest the centroid and plot its time series
plot(r_basin[[100]], main = "Regridded precip, one day", col = hcl.colors(50, "Blues"))
plot(basin_ll, add = TRUE, border = "red", lwd = 2)

ts_vec <- as.numeric(r_basin[cellFromXY(r_basin, crds(centroids(basin_ll)))])
plot(ts_vec, type = "l", xlab = "Day", ylab = "Precip (mm)", main = "One basin cell, all days")
sum(is.na(ts_vec))

# save so terra and the C program can both use it
# now has a real CRS and real 1-D lon/lat dims 
writeRaster(r_basin, file.path(basin_nc_dir, "wrf_prec_basin.nc"), overwrite = TRUE, filetype = "netcdf")
```

### convert and regrid temperature maximum data
```{r, tmax data}
# convert tmax data
tmax_dir <- "/Users/grace/rprojects/wrf_project/dly_tmax"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_tmax_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
tmax_files <- list.files(tmax_dir, pattern = "\\.nc$", full.names = TRUE)

# go over all files
all_tmax <- lapply(tmax_files, regrid_wrf_var, varname = "airtmax",
                    target_grid = target_grid, map_ix = map_ix, map_iy = map_iy,
                    is_cumulative = FALSE) # all variables other than precip are daily, not cumulative

r_basin <- rast(all_tmax) # concatenate all years along time
r_basin <- mask(r_basin, basin_ll) # drop cells outside the basin polygon

# sanity check
plot(r_basin[[100]], main = "Regridded air temperature max, one day", col = hcl.colors(50, "Blues"))
plot(basin_ll, add = TRUE, border = "red", lwd = 2)
ts_vec <- as.numeric(r_basin[cellFromXY(r_basin, crds(centroids(basin_ll)))])
plot(ts_vec, type = "l", xlab = "Day", ylab = "", main = "One basin cell, all days")
sum(is.na(ts_vec))

# save so terra and the C program can both use it
writeRaster(r_basin, file.path(basin_nc_dir, "wrf_tmax_basin.nc"), overwrite = TRUE, filetype = "netcdf")
```

### convert and regrid temperature minimum data
```{r, tmax data}
# convert tmax data
tmin_dir <- "/Users/grace/rprojects/wrf_project/dly_tmin"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_tmin_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
tmin_files <- list.files(tmin_dir, pattern = "\\.nc$", full.names = TRUE)

# go over all files
all_tmin <- lapply(tmin_files, regrid_wrf_var, varname = "airtmin",
                    target_grid = target_grid, map_ix = map_ix, map_iy = map_iy,
                    is_cumulative = FALSE) # all variables other than precip are daily, not cumulative

r_basin <- rast(all_tmin) # concatenate all years along time
r_basin <- mask(r_basin, basin_ll) # drop cells outside the basin polygon

# sanity check
plot(r_basin[[100]], main = "Regridded air temperature minimum, one day", col = hcl.colors(50, "Blues"))
plot(basin_ll, add = TRUE, border = "red", lwd = 2)

ts_vec <- as.numeric(r_basin[cellFromXY(r_basin, crds(centroids(basin_ll)))])
plot(ts_vec, type = "l", xlab = "Day", ylab = "", main = "One basin cell, all days")

# save so terra and the C program can both use it
writeRaster(r_basin, file.path(basin_nc_dir, "wrf_tmin_basin.nc"), overwrite = TRUE, filetype = "netcdf")
```

### convert and regrid relative humidity data
```{r, relhum data}
# convert tmax data
rh_dir <- "/Users/grace/rprojects/wrf_project/dly_relhum"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_relhum_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
rh_files <- list.files(rh_dir, pattern = "\\.nc$", full.names = TRUE)

# go over all files
all_rh <- lapply(rh_files, regrid_wrf_var, varname = "rhmean",
                    target_grid = target_grid, map_ix = map_ix, map_iy = map_iy,
                    is_cumulative = FALSE) # all variables other than precip are daily, not cumulative

r_basin <- rast(all_rh) # concatenate all years along time
r_basin <- mask(r_basin, basin_ll) # drop cells outside the basin polygon

# sanity check
plot(r_basin[[100]], main = "Regridded relative humidity, one day", col = hcl.colors(50, "Blues"))
plot(basin_ll, add = TRUE, border = "red", lwd = 2)
ts_vec <- as.numeric(r_basin[cellFromXY(r_basin, crds(centroids(basin_ll)))])
plot(ts_vec, type = "l", xlab = "Day", ylab = "", main = "One basin cell, all days")

# save so terra and the C program can both use it
writeRaster(r_basin, file.path(basin_nc_dir, "wrf_relhum_basin.nc"), overwrite = TRUE, filetype = "netcdf")
```

### convert and regrid maximum wind speed data
```{r, windspeed data}
# convert windspeed data
maxwind_dir <- "/Users/grace/rprojects/wrf_project/dly_winds"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_winds_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
windmax_files <- list.files(maxwind_dir, pattern = "\\.nc$", full.names = TRUE)

# go over all files
all_windmax <- lapply(windmax_files, regrid_wrf_var, varname = "u10max",
                    target_grid = target_grid, map_ix = map_ix, map_iy = map_iy,
                    is_cumulative = FALSE) # all variables other than precip are daily, not cumulative

r_basin <- rast(all_windmax) # concatenate all years along time
r_basin <- mask(r_basin, basin_ll) # drop cells outside the basin polygon

# sanity check
plot(r_basin[[100]], main = "Regridded maximum daily wind speed, one day", col = hcl.colors(50, "Blues"))
plot(basin_ll, add = TRUE, border = "red", lwd = 2)
ts_vec <- as.numeric(r_basin[cellFromXY(r_basin, crds(centroids(basin_ll)))])
plot(ts_vec, type = "l", xlab = "Day", ylab = "", main = "One basin cell, all days")
sum(is.na(ts_vec))

# save so terra and the C program can both use it
writeRaster(r_basin, file.path(basin_nc_dir, "wrf_winds_basin.nc"), overwrite = TRUE, filetype = "netcdf")
```

## Make the base station from cleaned .nc files
Make the base station itself - C code and R script mainly from Will
```{r, make base station}
# Requires
#   - createbaseinfo_netcdf.c compiled binary
# 
# Inputs:
#   - Processed/edited/subset aggregated (through time), netcdf climate data, processed in chunk above
#   - Basin and DEM raster maps

plots = T

# ------------------------------ INPUTS ------------------------------
# uncompiled source of the createbaseinfo_netcdf.c code
ncbase_src = "/Users/grace/rprojects/wrf_project/createbaseinfo_netcdf.c"

# USE SAME BASIN RASTER AS PRIOR
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire") 
DEM = rast("/Users/grace/rprojects/wrf_project/SWdem30m.tif") #digital elevation model

# location for maps to be used with base station creation
map_dest = "/Users/grace/rprojects/wrf_project/netcdfmaps"

# where to output a new zone map based on Netcdf grid
zone_dest = "/Users/grace/rprojects/wrf_project/preprocessing_newzone"

# specify climate files individually or via pattern, USE PROCESSED CLIMATE INPUTS FROM PREVIOUS SCRIPT 
clim_files_final = list.files(path = "/Users/grace/rprojects/wrf_project/merged_files", pattern = "\\.nc$", full.names = T)

# ------------------------------ END INPUTS ------------------------------

# for later use
basin_vect = as.polygons(basin)
basin_vect_latlon = project(basin_vect, "EPSG:4326")

# ------------------------------ GET GRID MAP FROM NETCDF ------------------------------
# there are slight differences between getting the grid from the metadata elevation map and the 
# netcdf already clipped using NCO. Using the clipped data from NCO as a default for now, but 
# could easily be better the other way, code should be easy to toggle

# commentary above is from will and I am just going to use the grid from my clipped climate net cdf files

# ---------- Grid from clipped climate netcdf files ----------
grid_from_clim = T
if (grid_from_clim) {
  # read in netcdf as raster, get only 1 day of precip
  nc_grid_lonlat = rast(clim_files_final[1])[[1]]
  # set values to sequential, change name
  values(nc_grid_lonlat) = seq_along(values(nc_grid_lonlat))
  names(nc_grid_lonlat) = "ID"
  
  # project to utm, mask to get zone/basestation map
  nc_grid = project(nc_grid_lonlat, basin, method="near") #will's original line
  
  # resample basin back to new grid (so extents align for masking) -- this didn't work
  #basin_resampled = resample(basin, nc_grid, method="near")
  
  # mask
  id_grid = mask(nc_grid, basin)
}

# --------------- output the zone grid for use in preprocessing ---------------
# maps need to be based on the netcdf latlong grid resolution + extent
#zone = mask(x = id_grid_proj, mask = basin) # will had x=id_grid_proj, replacing with id_grid didn't work
#writeRaster(id_grid, file.path(zone_dest, "zone_ids.tif"), overwrite=T)

# ------------------------------ GENERATE INPUTS FOR NETCDF BASE STATION CREATION  ------------------------------
# maps need to be based on the netcdf latlong grid resolution + extent
demproj = project(DEM, "EPSG:4326")
dem_nc = resample(demproj, nc_grid_lonlat, method="average")
# LAI
lai_nc = dem_nc
values(lai_nc) = 3.0
names(lai_nc) = "lai"

if (!file.exists(map_dest)) {
  dir.create(map_dest)
}

writeRaster(dem_nc, filename = file.path(map_dest,"dem.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(lai_nc, filename = file.path(map_dest,"lai.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(nc_grid_lonlat, filename = file.path(map_dest,"cellid.asc"), filetype="AAIGrid", datatype =  "INT4S", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)

file.remove(list.files(path = map_dest, pattern = ".prj|.aux.xml", full.names = T))

# format: ID ID Y X X Y

xyid_loc = cbind(1:nrow(crds(nc_grid_lonlat)), 1:nrow(crds(nc_grid_lonlat)), crds(nc_grid_lonlat)[,c("y","x")], crds(nc_grid_lonlat))

write.table(xyid_loc, file.path(map_dest,"xyid_loc.txt"), row.names = F, col.names = F)

# ------------------------------ RUN C BIN TO GET BASE STATION  ------------------------------
# COMPILE
system(noquote(paste("bash -c \"", paste0("gcc ", ncbase_src, " -o ", file.path(map_dest,"create_netcdfbase")), "\"", sep = "")))

# RUN -- YOU MAY NEED TO EDIT THIS, command should look like: ./create_netcdfbase cellid.asc lai.asc dem.asc xyid_loc.txt clim/ netcdf.base
cmd = paste0(file.path(map_dest,"create_netcdfbase"), " ", file.path(map_dest,"cellid.asc"), 
             " ", file.path(map_dest,"lai.asc"), " ",
             file.path(map_dest,"dem.asc"), " ", file.path(map_dest,"xyid_loc.txt"), 
             " ", dirname(clim_files_final[1]), "/ ", file.path(map_dest,"netcdf.base"))

system(noquote(paste("bash -c \"",cmd ,"\"", sep = "")))
```
