---
title: "sedgwick net cdf processing"
output: html_document
date: "2024-01-18"
---

## libraries
```{r, warning=FALSE, results=FALSE}
library(ncdf4)
library(raster)
library(sp)
library(sf)
library(maps)
library(terra)
library(sp)
#library(ncdf4.helpers)
library(ggplot2)
library(tidyverse)
```

## Logic checks
What is actually in each .nc file? Information:
   3 variables (excluding dimension variables):
      float XLAT[west_east,south_north]   (Contiguous storage)  
          _FillValue: NaN
          FieldType: 104
          MemoryOrder: XY 
          description: LATITUDE, SOUTH IS NEGATIVE
          units: degree_north
          stagger: 
          coordinates: XLONG XLAT
      float XLONG[west_east,south_north]   (Contiguous storage)  
          _FillValue: NaN
          FieldType: 104
          MemoryOrder: XY 
          description: LONGITUDE, WEST IS NEGATIVE
          units: degree_east
          stagger: 
          coordinates: XLONG XLAT
      double precsum[west_east,south_north,times]   (Contiguous storage)  
          _FillValue: NaN
          precsum: daily total precipitation K)
          coordinates: XLONG XLAT
   3 dimensions:
      times  Size:366 
          units: days since 1987-07-01 00:00:00
          calendar: proleptic_gregorian
      south_north  Size:234 (no dimvar)
      west_east  Size:258 (no dimvar)

What does this tell us? 
- Data stored in 3D grid: x-direction (longitude, west_east dimension here), y-direction (latitude, south_north dimension here), time (daily)
- 3 data variables (excluding dimension)
- Model has 258 × 234 spatial points
- Each value in times is a day offset from 1987-07-01 00:00:00 (standard Gregorian)

- There are 3 data variables (excluding dimensions): XLAT, XLONG, and specific climate variable (precsum, relhum, etc.)
- XLAT: degree_north, float, 2D grid [west_east, south_north], float data type
- XLONG: degree_east (west is negative, not sure what this means), 2D grid [west_east,south_north], float data type
- WRF grids are curvilinear (curvilinear orthogonal coordinates (often referred to as Arakawa C-grid or map projections) are used to represent the Earth's spherical surface on a 2D computational plane) 
- Each grid cell has its own latitude and longitude
- Together, XLAT + XLONG define the spatial location of every grid cell

- Dimensions are ordered as [west_east, south_north, time] (more broadly: [x, y, time])
- First index → longitude direction; second index → latitude direction; third index → time
- For one grid cell, a daily time series is (example: precip): precsum[i, j, ] (to connect to geography: lat  <- XLAT[i, j], lon  <- XLONG[i, j])

What is actually in the SWbasin_AvgFire file?
- Datum: NAD83
- Projection: UTM Zone 10 North
- Units: meters
- EPSG code: 26910 (totally reasonable for a basin raster (engineering/hydrology workflows often use UTM)
- NOTE: WRF grid points live in EPSG:4326 (lon/lat) (must reproject the basin polygon)
- WRF grids are curvilinear — projecting them introduces unnecessary distortion and complexity. Always project the basin to lon/lat instead

Notes from debugging:
- wrf_sub <- wrf_vect[basin_vect, ] does not work: SpatVector[i, j] is interpreted as attribute subsetting, not spatial, so basin_ll is being treated as an invalid row index. Result: an empty object with attribute names (ix, iy) and NAs. This is `terra()` specific, and is different from sf
- x[y] performs spatial subsetting, intersect(x, y) gives us spatial intersection
- basin_vect_proj <- project(basin_ll, crs(nc_example)) gives the following Error: [project] output crs is not valid. rast(nc_path) reads the raster from the NetCDF, while WRF NetCDF rasters often have a non-standard CRS internally. Terra sees it as a CRS, but it may not be a PROJ string that project() can handle. Even though the raster prints a CRS, terra cannot parse it for project()
- Numbers from WRF data are usually grid indices, not degrees: an invalid NetCDF georeferencing situation. `terra()` thinks that the numbers it receives are latitude and longitude, but they're actually row and column indices. 

| Object           | Lives in                                     |
| ---------------- | -------------------------------------------- |
| `wrf_sub$ix, iy` | **WRF index space**                          |
| `r`              | **WRF index space (mis-labeled as lon/lat)** |
| `XLAT / XLONG`   | **True geographic space**                    |
| `basin_ll`       | **True geographic space**                    |


# MOST RECENT CORRECT ATTEMPT BELOW:
```{r}
# define and reproject Sedgwick basin
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")
basin_vect = as.polygons(basin, dissolve = TRUE, na.rm = TRUE)
basin_ll <- project(basin_vect, "EPSG:4326")

# create WRF grid points from XLON/XLONG -- this is metadata for where wrf grid points are
## choose example file
nc <- nc_open("/Users/grace/rprojects/wrf_project/dly_prec/wrf.dlytotal.precip.1988.nc")
lon <- ncvar_get(nc, "XLONG")
lat <- ncvar_get(nc, "XLAT")
dims <- dim(lon) 

## flatten and index
wrf_pts <- data.frame(lon = as.vector(lon), lat = as.vector(lat),
  ix  = rep(seq_len(dims[1]), times = dims[2]), iy  = rep(seq_len(dims[2]), each  = dims[1]))
head(wrf_pts)

## convert to terra spatial vector (SpatVect)
wrf_vect <- vect(wrf_pts, geom = c("lon", "lat"), crs = "EPSG:4326")

## crop using converted basin vector and with terra specific formatting
wrf_sub <- wrf_vect[basin_ll] #contains only spatial metadata
wrf_sub

## sanity check for this step:
# rel <- relate(wrf_vect, basin_ll, "intersects")
# sum(rel) #109 overlaps

## sanity checks: visual
plot(wrf_sub, pch = ".", cex = 0.3)
plot(basin_ll, add = TRUE, border = "red", lwd = 2)
```

Process climate data
```{r, process precip data}
# loop over all precipitation information
prec_dir <- "/Users/grace/rprojects/wrf_project/dly_prec"
basin_nc_dir  <- "/Users/grace/rprojects/wrf_project/dly_prec_basin"
if(!dir.exists(basin_nc_dir)) dir.create(basin_nc_dir)
prec_files <- list.files(prec_dir, pattern = "\\.nc$", full.names = TRUE)

# Load the basin subset metadata (grid indices)
# wrf_sub must have ix, iy columns
head(wrf_sub)

process_prec_file <- function(nc_file, wrf_sub) {
  # Read the raster
  r <- rast(nc_file)
  
  n_days <- nlyr(r)
  n_cells <- nrow(wrf_sub)
  
  # Pre-allocate output matrix: rows = grid cells, columns = days
  basin_vals <- matrix(NA, nrow = n_cells, ncol = n_days)
  
  # Extract values using WRF grid indices
  for (i in seq_len(n_cells)) {
    ix <- wrf_sub$ix[i]
    iy <- wrf_sub$iy[i]
    basin_vals[i, ] <- as.numeric(r[iy, ix, ])
  }
  
  # Convert cumulative -> daily
  basin_vals_daily <- t(apply(basin_vals, 1, function(x) c(x[1], diff(x))))
  
  # Convert mm -> m
  basin_vals_daily <- basin_vals_daily / 1000
  
  return(basin_vals_daily)
}

all_prec <- lapply(prec_files, process_prec_file, wrf_sub = wrf_sub)
vals_combined <- do.call(cbind, all_prec) #combine all years along the time dimension (columns)
stopifnot(!is.null(vals_combined))
stopifnot(nrow(vals_combined) == nrow(wrf_sub))

# sanity checks
# Example with first cell
ts_vec <- process_prec_file(prec_files[1], wrf_sub)[1, ]
summary(ts_vec)
plot(ts_vec, type = "l",
     main = "Daily precipitation at first WRF basin cell",
     xlab = "Day", ylab = "Precipitation (m)")

# Quick check: number of NAs
sum(is.na(ts_vec))

# Use terra to create SpatRaster with proper dimensions
n_cells <- nrow(wrf_sub)
n_days <- ncol(vals_combined)

# Create a raster stack with one layer per day
ras_list <- lapply(1:n_days, function(day) {
  # Convert column to matrix (rows x cols = 1 row per cell, 1 col)
  mat <- matrix(vals_combined[, day], nrow = n_cells, ncol = 1)
  rast(mat)
})

r_basin <- rast(ras_list)

# Assign names
names(r_basin) <- paste0("day_", 1:n_days)

# Save
writeRaster(r_basin, "/Users/grace/rprojects/wrf_project/dly_prec_basin/wrf_prec_basin.nc",
            overwrite = TRUE, filetype = "netcdf")
```

```{r}
# sanity checks for precip conversion overall
## sanity check for location and convention
r_check <- rast(merged_file)
r_check

## visual sanity check: single wrf cell (if that looks spiky and non-cumulative, you’re golden)
## first convert
r <- rast("/Users/grace/rprojects/wrf_project/dly_prec_basin/wrf.dlytotal.precip.1988.nc")
r
nlyr(r)

# Pick a single WRF cell from your basin subset
ix <- wrf_sub$ix[1]  # column index in WRF grid
iy <- wrf_sub$iy[1]  # row index in WRF grid

# Extract the time series for this cell
# Use row/column indexing (index-space, NOT cell numbers)
ts_vec <- as.numeric(r[iy, ix, ])  # extract all layers for that cell

# Sanity check summaries
summary(ts_vec)
range(ts_vec, na.rm = TRUE)

# Plot safely (guard against all NA)
# ------------------------------
if (all(is.na(ts_vec))) {
  message("Warning: all values are NA. Check WRF indices or raster layers.")
} else {
  plot(ts_vec, type = "l",
       xlab = "Day",
       ylab = "Daily precipitation (m)",
       main = paste("Daily precipitation at WRF cell (ix=", ix, ", iy=", iy, ")"))
}

## second visual sanity check, load merged basin NetCDF
r <- rast("/Users/grace/rprojects/wrf_project/dly_prec_basin/wrf_precip_daily_basin_all_years.nc")
crs(r) <- "EPSG:4326"
day <- 100  #choose a day with precipitation (skip day 1 if it's often zero)
plot(r[[day]],
     main = paste("WRF daily precipitation – day", day),
     col = hcl.colors(50, "Blues"))

plot(basin_ll, add = TRUE, border = "red", lwd = 2)

## another sanity check: WRF points versus raster cells (check that you did not lose or shift WRF grid cells)
plot(r[[day]],
     main = "WRF grid alignment check",
     col = hcl.colors(50, "Blues"))

plot(wrf_sub, add = TRUE, col = "black", pch = 16, cex = 0.4)
plot(basin_ll, add = TRUE, border = "red", lwd = 2)

## sanity check: distribution snapshot (spatial statistics, do values on this day look physically reasonable?)
vals <- values(r[[day]])
vals <- vals[!is.na(vals)]

summary(vals)
hist(vals, breaks = 40,
     main = paste("Daily precipitation distribution – day", day),
     xlab = "m/day")

```

```{r, check extraction -- for processing a single index of precip data}
nc <- nc_open("/Users/grace/rprojects/wrf_project/dly_prec/wrf.dlytotal.precip.1988.nc")
time <- ncvar_get(nc, "times")
time_units <- ncatt_get(nc, "times", "units")$value
time_origin <- sub("days since ", "", time_units)
dates <- as.Date(time, origin = time_origin)
head(dates)
tail(dates)
length(dates)

i <- wrf_sub$ix[1]
j <- wrf_sub$iy[1]

prec <- ncvar_get(
  nc,
  "precsum",
  start = c(i, j, 1),
  count = c(1, 1, -1)
)
length(prec)       
summary(prec)
plot(dates, prec, type = "l")
```

## Make the base station from cleaned .nc files
Make the base station itself - C code and R script mainly from Will
```{r, make base station}
# Requires
#   - createbaseinfo_netcdf.c compiled binary
# 
# Inputs:
#   - Processed/edited/subset aggregated (through time), netcdf climate data, processed in chunk above
#   - Basin and DEM raster maps

plots = T

# ------------------------------ INPUTS ------------------------------
# uncompiled source of the createbaseinfo_netcdf.c code
ncbase_src = "/Users/grace/rprojects/wrf_project/createbaseinfo_netcdf.c"

# USE SAME BASIN RASTER AS PRIOR
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire") 
DEM = rast("/Users/grace/rprojects/wrf_project/SWdem30m.tif") #digital elevation model

# location for maps to be used with base station creation
map_dest = "/Users/grace/rprojects/wrf_project/netcdfmaps"

# where to output a new zone map based on Netcdf grid
zone_dest = "/Users/grace/rprojects/wrf_project/preprocessing_newzone"

# specify climate files individually or via pattern, USE PROCESSED CLIMATE INPUTS FROM PREVIOUS SCRIPT 
clim_files_final = list.files(path = "/Users/grace/rprojects/wrf_project/merged_files", pattern = "\\.nc$", full.names = T)

# ------------------------------ END INPUTS ------------------------------

# for later use
basin_vect = as.polygons(basin)
basin_vect_latlon = project(basin_vect, "EPSG:4326")

# ------------------------------ GET GRID MAP FROM NETCDF ------------------------------
# there are slight differences between getting the grid from the metadata elevation map and the 
# netcdf already clipped using NCO. Using the clipped data from NCO as a default for now, but 
# could easily be better the other way, code should be easy to toggle

# commentary above is from will and I am just going to use the grid from my clipped climate net cdf files

# ---------- Grid from clipped climate netcdf files ----------
grid_from_clim = T
if (grid_from_clim) {
  # read in netcdf as raster, get only 1 day of precip
  nc_grid_lonlat = rast(clim_files_final[1])[[1]]
  # set values to sequential, change name
  values(nc_grid_lonlat) = seq_along(values(nc_grid_lonlat))
  names(nc_grid_lonlat) = "ID"
  
  # project to utm, mask to get zone/basestation map
  nc_grid = project(nc_grid_lonlat, basin, method="near") #will's original line
  
  # resample basin back to new grid (so extents align for masking) -- this didn't work
  #basin_resampled = resample(basin, nc_grid, method="near")
  
  # mask
  id_grid = mask(nc_grid, basin)
}

# --------------- output the zone grid for use in preprocessing ---------------
# maps need to be based on the netcdf latlong grid resolution + extent
zone = mask(x = id_grid_proj, mask = basin) # will had x=id_grid_proj, replacing with id_grid didn't work
writeRaster(id_grid, file.path(zone_dest, "zone_ids.tif"), overwrite=T)

# ------------------------------ GENERATE INPUTS FOR NETCDF BASE STATION CREATION  ------------------------------
# maps need to be based on the netcdf latlong grid resolution + extent
demproj = project(DEM, "EPSG:4326")
dem_nc = resample(demproj, nc_grid_lonlat, method="average")
# LAI
lai_nc = dem_nc
values(lai_nc) = 3.0
names(lai_nc) = "lai"

if (!file.exists(map_dest)) {
  dir.create(map_dest)
}

writeRaster(dem_nc, filename = file.path(map_dest,"dem.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(lai_nc, filename = file.path(map_dest,"lai.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(nc_grid_lonlat, filename = file.path(map_dest,"cellid.asc"), filetype="AAIGrid", datatype =  "INT4S", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
file.remove(list.files(path = map_dest, pattern = ".prj|.aux.xml", full.names = T))

# format: ID ID Y X X Y

xyid_loc = cbind(1:nrow(crds(nc_grid_lonlat)), 1:nrow(crds(nc_grid_lonlat)), crds(nc_grid_lonlat)[,c("y","x")], crds(nc_grid_lonlat))
write.table(xyid_loc, file.path(map_dest,"xyid_loc.txt"), row.names = F, col.names = F)

# ------------------------------ RUN C BIN TO GET BASE STATION  ------------------------------
# COMPILE
system(noquote(paste("bash -c \"", paste0("gcc ", ncbase_src, " -o ", file.path(map_dest,"create_netcdfbase")), "\"", sep = "")))

# RUN -- YOU MAY NEED TO EDIT THIS, command should look like: ./create_netcdfbase cellid.asc lai.asc dem.asc xyid_loc.txt clim/ netcdf.base
cmd = paste0(file.path(map_dest,"create_netcdfbase"), " ", file.path(map_dest,"cellid.asc"), " ", file.path(map_dest,"lai.asc"), " ",
             file.path(map_dest,"dem.asc"), " ", file.path(map_dest,"xyid_loc.txt"), " ", dirname(clim_files_final[1]), "/ ", file.path(map_dest,"netcdf.base"))
system(noquote(paste("bash -c \"",cmd ,"\"", sep = "")))
```


# ORIGINAL ATTEMPT BELOW:

## extract GPS coordinates from WRF data 
Code chunk shows the extent of latitude and longitude for WRF data from Charles, and compares it to the spatial extent of Sedgwick. 
```{r}
# extract coordinates for overall WRF data
nc_file <- nc_open("/Users/grace/rprojects/wrf_project/dly_prec/wrf.dlytotal.precip.1988.nc")
nc_file

latitude <- ncvar_get(nc_file, "XLAT")
longitude <- ncvar_get(nc_file, "XLONG")
nc_close

# put extracted coordinates in a df
longlat_df <- data.frame(longitude = as.vector(longitude), latitude=as.vector(latitude))
#crs(longlat_df) <- "+proj=longlat +datum=WGS84"
california_map <- ggplot2::map_data("state", region = "california") #for base map

# plot coordinates on a map for overall WRF data
ggplot() + geom_polygon(data = california_map, aes(x = long, y = lat, group = group), fill = "grey", color = "black") +
  geom_point(data = longlat_df, aes(x = longitude, y = latitude), color = "darkred", size = 1) +
  theme_bw() + labs(x = "Longitude", y = "Latitude", title = "Spatial Extent of Coordinates for Original File")

# extract coordinates for cropped
nc_file <- nc_open("/Users/grace/rprojects/wrf_project/crop_files/crop_wrf.dlytotal.precip.1988.nc")
latitude <- ncvar_get(nc_file, "XLAT")
longitude <- ncvar_get(nc_file, "XLONG")
nc_close

# put extracted coordinates in a df
longlat_df <- data.frame(longitude = as.vector(longitude), latitude=as.vector(latitude))
#crs(longlat_df) <- "+proj=longlat +datum=WGS84"
california_map <- map_data("state", region = "california") #for base map


# plot coordinates on a map
ggplot() + geom_polygon(data = california_map, aes(x = long, y = lat, group = group), fill = "grey", color = "black") +
  geom_point(data = longlat_df, aes(x = longitude, y = latitude), color = "darkred", size = 1) +
  theme_bw() + labs(x = "Longitude", y = "Latitude", title = "Spatial Extent of Coordinates for Cropped File for Sedgwick Basin Extent")

# for seeing the Sedgwick basin
# get basin extent and plot
plot = T
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")
basin_vect = as.polygons(basin)

# projection for wrf netcdf data is: CHECK THIS, IS PROBSBLY DIFFERENT FROM WILL'S
basin_vect_unproj = project(basin_vect, "+proj=longlat +datum=WGS84 +no_defs ")
if (plot) {plot(basin_vect_unproj)}
basin_extent = ext(basin_vect_unproj)

# explicit data type checks for the raster I'm using
crs(basin)
ext(basin)
res(basin)
nlyr(basin)
crs(basin_vect)
```

## Crop netCDF data to the extent of the basin of the watershed

Create extent for basin
```{r, process grid met}
# give example climate files to iterate through - this is just for testing
#clim_files = list.files(path = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/tmax/test/",pattern = "\\.nc$", full.names = T)

# ------------------------------ CROP + PROCESS NETCDF CLIMATE DATA USING BASIN EXTENT AND NCO -----------------------------

# FOR NETCDF TO WORK WITH RHESSYS:
#   Must use 'time' as temporal dimension, below we change every instance of 'day' to 'time' to be safe
#   Must have variable as floating point/double data type, NOT INTEGER

basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")

basin_vect = as.polygons(basin)
basin_vect_latlon = project(basin_vect, "EPSG:4326")

basin_extent = ext(basin_vect_latlon)
# cell size is 4 km, buffer should be at least 3 km at most latitudes 
# so should cover the centroids of netcdf cells not completely inside basin
# Charles' WRF is 2km spatial resolution, so halving what Will did
basin_extent_buffer = extend(basin_extent, 0.015)

# this should be in lon lat wgs84 to match ncdf - see xy axes
if (plot) {
  plot(basin_extent_buffer)
  plot(basin_extent,add=T)
  plot(basin_vect_latlon,add=T)
}

basin_extent = basin_extent_buffer
```

Loop through the climate variables to crop (run four or five times depending on variables needed)
```{r, windspeed and crop}
# specify the paths for each of the 5 variable types
dly_prcp_path = "/Users/grace/rprojects/wrf_project/dly_prec"
dly_relhum_path = "/Users/grace/rprojects/wrf_project/dly_relhum"
#windmax_path = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/windmax"
tmax_path = "/Users/grace/rprojects/wrf_project/dly_tmax"
tmin_path = "/Users/grace/rprojects/wrf_project/dly_tmin"

# Destination folder for cropped files
clim_dest = "/Users/grace/rprojects/wrf_project/crop_files"

# process each climate variable type separately
# update this for each of the 5 variable types
clim_files = list.files(path = tmin_path, pattern = "\\.nc$", full.names = T)

# Loop through climate files
for (infile in clim_files) {
  
  outfile <- paste0(file.path(clim_dest, "crop_"), basename(infile))
  
  # Define the longitude and latitude bounds from the basin_extent (plus buffer) defined above
  lon_min <- xmin(basin_extent)
  lon_max <- xmax(basin_extent)
  lat_min <- ymin(basin_extent)
  lat_max <- ymax(basin_extent)
  
  # Build cropping command using cdo sellonlatbox based on basin_extent
  crop_cmd <- paste0("cdo sellonlatbox,", lon_min, ",", lon_max, ",", lat_min, ",", lat_max, " ", infile, " ", outfile)
  system(crop_cmd)
  
  # Check if the cropped file was successfully created
  if (!file.exists(outfile)) {
    cat("Error: Cropping failed for", infile, "\n")
    next  # Skip to the next file if cropping failed
  }
  
  cat("Successfully cropped:", infile, "->", outfile, "\n")
  
  # Open the cropped NetCDF file
  nc <- nc_open(outfile)
  
  # Get variable names
  var_names <- names(nc$var)
  nc_close(nc)
  
  # If there are no variables, skip processing
  if (length(var_names) == 0) {
    cat("Warning: No valid variables found in", infile, "\n")
    next
  }
  
  # Special handling for wind files (compute wind speed from U and V components)
  if (grepl("wind", infile) && all(c("U", "V") %in% var_names)) {
    nc <- nc_open(outfile, write = TRUE)
    
    u_component <- ncvar_get(nc, "U")
    v_component <- ncvar_get(nc, "V")
    wind_speed <- sqrt(u_component^2 + v_component^2)
    
    # Create a new variable in the NetCDF file for wind speed
    wind_def <- ncvar_def("wind_speed", "m/s", nc$dim)
    nc_outfile <- paste0(file.path(clim_dest, "wind_speed_"), basename(outfile))
    nc_create(nc_outfile, list(dim_def = nc$dim, var_def = wind_def))
    
    nc_out <- nc_open(nc_outfile, write = TRUE)
    ncvar_put(nc_out, "wind_speed", wind_speed)
    nc_close(nc_out)
    
    cat("Processed wind speed:", infile, "->", nc_outfile, "\n")
    nc_close(nc)
  }
  
  # Convert each variable in the NetCDF file to double precision
  for (var in var_names) {
    ncap_cmd <- paste0("ncap2 -O -s '", var, "=double(", var, ")' ", outfile, " ", outfile)
    system(ncap_cmd)
  }
  
  # Print confirmation
  cat("Processed:", infile, "->", outfile, "\n")
}
```

## Unit check prior to conversion
Precipitation: precsum, range 0-891.165; mm
Relative humidity: rhmean, range 4.707718-100; %
Wind speed max: maxspeed, range 3.833371-16.79659; idk
Temperature maximum: airtmax, range 258.0816-317.1882; K
Temperature minimum: airtmin, range 251.7519-305.2329; K
```{r, check variable names and units within wrf files}
# files that need to be converted, check
dly_prcp_file = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/dly_prec/wrf.dlytotal.precip.1988.nc"
dly_relhum_file = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/dly_relhum/wrf.dlymean.rh2m.1988.nc"
windmax_file = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/windmax/wrf.dmax.wind10m.1999-07-01.nc"
tmax_file = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/tmax/wrf.dlymax.airt2m.1988.nc"
tmin_file = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/tmin/wrf.dlymin.airt2m.1988.nc"

nc = nc_open(dly_relhum_file)
var_name = names(nc$var)[3]   # gets the variable name
var_name

var_values = ncvar_get(nc, var_name)
var_range = range(var_values, na.rm = TRUE) 
cat("min:", var_range[1], "max:", var_range[2], "\n")
nc_close(nc)

var_name = names(nc$var)[2]   # gets the variable name
var_name

names(nc$var)
```

Convert the units from netCDF files into needed format for RHESSys base station. 
```{r, convert units for cropped .nc files}
# Function to convert NetCDF data and save in a new folder
convert_nc_units <- function(nc_file, output_folder) {
  
  # Open NetCDF file
  nc <- nc_open(nc_file, write = FALSE)  # Open in read-only mode
  
  # Extract variable name
  var_name <- names(nc$var)[1]
  data <- ncvar_get(nc, var_name)
  
  # Define output file path in the new folder
  output_file <- file.path(output_folder, basename(nc_file))
  
  # Determine conversion based on file name
  if (grepl("dlymax.airt2m", nc_file)) {
    data <- data - 273.15  # Convert K to °C
    cat("Converted air temperature from K to °C:", nc_file, "\n")
  } else if (grepl("dlymin.airt2m", nc_file)) {
    data <- data - 273.15  # Convert K to °C
    cat("Converted air temperature from K to °C:", nc_file, "\n")
  } else if (grepl("dlytotal.precip", nc_file)) {
    data <- data / 1000  # Convert mm to m
    cat("Converted precipitation from mm to m:", nc_file, "\n")
  } else if (grepl("dmax.wind10m", nc_file)) {
    cat("Wind data does not require conversion:", nc_file, "\n")
  }
  
  # Close original NetCDF file
  nc_close(nc)
  
  # Copy original file to the new directory before modification
  file.copy(nc_file, output_file, overwrite = TRUE)
  
  # Reopen copied NetCDF file in write mode to modify it
  nc_out <- nc_open(output_file, write = TRUE)
  
  # Update variable with converted data
  ncvar_put(nc_out, var_name, data)
  
  # Close modified NetCDF file
  nc_close(nc_out)
  
  cat("Saved converted file to:", output_file, "\n")
}

# Define input and output paths
crop_path = "/Users/grace/rprojects/wrf_project/crop_files"
output_path = "/Users/grace/rprojects/wrf_project/crop_files_converted"

# Ensure output folder exists
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
}

# Get list of NetCDF files to process
nc_files = list.files(path = crop_path, pattern = "\\.nc$", full.names = TRUE)

# loop through files and convert
for (file in nc_files) {
  convert_nc_units(file, output_path)
}
```

## Upscale data from 2m resolution to 30m resolution - not working ?
```{r}
# set the factor needed, in this case we need a factor of 45 as the original resolution is 2m and the desired resolution is 90m
target_res = 30

# set input files and new output directory
input_path <- "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/crop_files_converted"
output_path <- "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/aggregated_files1"

# ensure output folder exists
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
}

# get list of NetCDF files to process
nc_files = list.files(path = input_path, pattern = "\\.nc$", full.names = TRUE)

# Loop through files
for (file_path in nc_files) {
  message("Processing: ", basename(file_path))
  
  # Load file
  r <- rast(file_path)
  
  # Get original resolution (assumes square cells)
  orig_res <- res(r)[1]
  
  # Calculate aggregation factor (round to nearest integer)
  agg_factor <- round(target_res / orig_res)
  
  if (agg_factor < 1) {
    message("File already coarser than target resolution — skipping aggregation.")
    r_agg <- r
  } else {
    # Aggregate spatially
    r_agg <- aggregate(r, fact = agg_factor, fun = mean, na.rm = TRUE)
  }
  
  # Save new file
  out_file <- file.path(output_path, paste0("agg", target_res, "m_", basename(file_path)))
  writeCDF(r_agg, out_file, overwrite = TRUE)
  
  message("Saved: ", out_file)
}
```

Debugging aggregation - resolution seems to be working?
```{r}
# checking against original script that just aggregated automatically to 90 and didn't work super great
f = "/Volumes/mambo_hd/Sedgwick/wrf/netcdf/crop_files_converted/crop_wrf.dlymax.airt2m.1988.nc"
r = rast(f)

res(r)
ext(r)
ncell(r)
crs(r)

# Load original and aggregated
r_orig <- rast("/Volumes/mambo_hd/Sedgwick/wrf/netcdf/crop_files_converted/crop_wrf.dlymax.airt2m.1988.nc")
r_agg  <- rast("/Volumes/mambo_hd/Sedgwick/wrf/netcdf/aggregated_files/agg90m_crop_wrf.dlymax.airt2m.1988.nc")

# Get centroids of each cell (just for the first layer to simplify)
centroids_orig <- as.points(r_orig[[1]])
centroids_agg  <- as.points(r_agg[[1]])

# Plot comparison
plot(ext(r_orig), col = "lightgrey", main = "Cell Centroids Comparison")
points(centroids_orig, pch = 1, col = "blue", cex = 0.5)
points(centroids_agg, pch = 3, col = "red", cex = 0.7)
legend("topright", legend = c("Original (Blue)", "Aggregated (Red)"), pch = c(1,3))
```

Debugging -- centroid and value output sanity check
```{r}
# File path
# checking a correctly cropped and converted value to see if that is the spatial issue
nc_path <- "/Users/grace/rprojects/wrf_project/crop_files/crop_wrf.dlytotal.precip.1988.nc"

# Read NetCDF as a SpatRaster
r <- rast(nc_path)

# If multiple time steps, calculate mean precipitation over the year
r_mean <- mean(r, na.rm = TRUE)

# Convert raster to data frame with coordinates
df <- as.data.frame(r_mean, xy = TRUE)
colnames(df) <- c("lon", "lat", "precip")

# Remove NA values
df <- df %>% filter(!is.na(precip))

# Plot 1: centroids only
p1 <- ggplot(df, aes(x = lon, y = lat)) +
  geom_point(size = 0.8, color = "blue") +
  coord_equal() +
  labs(title = "Centroids of Aggregated NetCDF Grid",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

# Plot 2: climate variable distribution
p2 = ggplot(df, aes(x = lon, y = lat, color = precip)) +
  geom_point(size = 1) +
  scale_color_viridis_c(name = "clim var", option = "plasma") +
  coord_equal() +
  labs(title = "Climate Variable Distribution Example",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

# Print plots
print(p1)
print(p2)
```

## Consolidate into single .nc files
```{r}
# takes originally converted files -- trying something new
input_dir <- "/Users/grace/rprojects/wrf_project/crop_files_converted"

# Output folder to write merged .nc files
output_dir <- "/Users/grace/rprojects/wrf_project/merged_files"

# ensure output folder exists
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Define the variable types you expect to process
vars <- c("dlymax.airt2m", "dlymin.airt2m", "dlytotal.precip", "dlymean.rh2m") #, "dmax.wind10m")  

# ---- Merge loop ----

for (v in vars) {
  message("Merging variable: ", v)
 
  # Get list of files for that variable
  files <- list.files(input_dir, pattern = paste0(v, ".*\\.nc$"), full.names = TRUE)
 
  # Sort by year using the number in the filename (assumes 4-digit year)
  files <- files[order(as.numeric(stringr::str_extract(files, "\\d{4}")))]
 
  # Read and stack all rasters
  r_list <- lapply(files, rast)
  r_stack <- rast(r_list)  # creates multilayer time stack
 
  # Output file name
  out_file <- file.path(output_dir, paste0(v, "_final.nc"))
 
  # Write to NetCDF
  writeCDF(r_stack, out_file, overwrite = TRUE)
 
  message("Saved to: ", out_file)
}
```

## Make the base station from cleaned .nc files
Make the base station itself - C code and R script mainly from Will
```{r, make base station}
# Requires
#   - createbaseinfo_netcdf.c compiled binary
# 
# Inputs:
#   - Processed/edited/subset aggregated (through time), netcdf climate data, processed in chunk above
#   - Basin and DEM raster maps

plots = T

# ------------------------------ INPUTS ------------------------------
# uncompiled source of the createbaseinfo_netcdf.c code
ncbase_src = "/Users/grace/rprojects/wrf_project/createbaseinfo_netcdf.c"

# USE SAME BASIN RASTER AS PRIOR
basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire") 
DEM = rast("/Users/grace/rprojects/wrf_project/SWdem30m.tif") #digital elevation model

# location for maps to be used with base station creation
map_dest = "/Users/grace/rprojects/wrf_project/netcdfmaps"

# where to output a new zone map based on Netcdf grid
zone_dest = "/Users/grace/rprojects/wrf_project/preprocessing_newzone"

# specify climate files individually or via pattern, USE PROCESSED CLIMATE INPUTS FROM PREVIOUS SCRIPT 
clim_files_final = list.files(path = "/Users/grace/rprojects/wrf_project/merged_files", pattern = "\\.nc$", full.names = T)

# ------------------------------ END INPUTS ------------------------------

# for later use
basin_vect = as.polygons(basin)
basin_vect_latlon = project(basin_vect, "EPSG:4326")

# ------------------------------ GET GRID MAP FROM NETCDF ------------------------------
# there are slight differences between getting the grid from the metadata elevation map and the 
# netcdf already clipped using NCO. Using the clipped data from NCO as a default for now, but 
# could easily be better the other way, code should be easy to toggle

# commentary above is from will and I am just going to use the grid from my clipped climate net cdf files

# ---------- Grid from clipped climate netcdf files ----------
grid_from_clim = T
if (grid_from_clim) {
  # read in netcdf as raster, get only 1 day of precip
  nc_grid_lonlat = rast(clim_files_final[1])[[1]]
  # set values to sequential, change name
  values(nc_grid_lonlat) = seq_along(values(nc_grid_lonlat))
  names(nc_grid_lonlat) = "ID"
  
  # project to utm, mask to get zone/basestation map
  nc_grid = project(nc_grid_lonlat, basin, method="near") #will's original line
  
  # resample basin back to new grid (so extents align for masking) -- this didn't work
  #basin_resampled = resample(basin, nc_grid, method="near")
  
  # mask
  id_grid = mask(nc_grid, basin)
}

# --------------- output the zone grid for use in preprocessing ---------------
# maps need to be based on the netcdf latlong grid resolution + extent
zone = mask(x = id_grid_proj, mask = basin) # will had x=id_grid_proj, replacing with id_grid didn't work
writeRaster(id_grid, file.path(zone_dest, "zone_ids.tif"), overwrite=T)

# ------------------------------ GENERATE INPUTS FOR NETCDF BASE STATION CREATION  ------------------------------
# maps need to be based on the netcdf latlong grid resolution + extent
demproj = project(DEM, "EPSG:4326")
dem_nc = resample(demproj, nc_grid_lonlat, method="average")
# LAI
lai_nc = dem_nc
values(lai_nc) = 3.0
names(lai_nc) = "lai"

if (!file.exists(map_dest)) {
  dir.create(map_dest)
}

writeRaster(dem_nc, filename = file.path(map_dest,"dem.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(lai_nc, filename = file.path(map_dest,"lai.asc"), filetype="AAIGrid", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
writeRaster(nc_grid_lonlat, filename = file.path(map_dest,"cellid.asc"), filetype="AAIGrid", datatype =  "INT4S", gdal = c("FORCE_CELLSIZE=TRUE"), NAflag=-9999, overwrite=T)
file.remove(list.files(path = map_dest, pattern = ".prj|.aux.xml", full.names = T))

# format: ID ID Y X X Y

xyid_loc = cbind(1:nrow(crds(nc_grid_lonlat)), 1:nrow(crds(nc_grid_lonlat)), crds(nc_grid_lonlat)[,c("y","x")], crds(nc_grid_lonlat))
write.table(xyid_loc, file.path(map_dest,"xyid_loc.txt"), row.names = F, col.names = F)

# ------------------------------ RUN C BIN TO GET BASE STATION  ------------------------------
# COMPILE
system(noquote(paste("bash -c \"", paste0("gcc ", ncbase_src, " -o ", file.path(map_dest,"create_netcdfbase")), "\"", sep = "")))

# RUN -- YOU MAY NEED TO EDIT THIS, command should look like: ./create_netcdfbase cellid.asc lai.asc dem.asc xyid_loc.txt clim/ netcdf.base
cmd = paste0(file.path(map_dest,"create_netcdfbase"), " ", file.path(map_dest,"cellid.asc"), " ", file.path(map_dest,"lai.asc"), " ",
             file.path(map_dest,"dem.asc"), " ", file.path(map_dest,"xyid_loc.txt"), " ", dirname(clim_files_final[1]), "/ ", file.path(map_dest,"netcdf.base"))
system(noquote(paste("bash -c \"",cmd ,"\"", sep = "")))
```

Chat GPT using DEM to create base station map
```{r}
# load in what you just created
dem = raster("/Users/grace/rprojects/wrf_project/netcdfmaps_final/dem.asc")
lai = raster("/Users/grace/rprojects/wrf_project/netcdfmaps_final/lai.asc")
cellid = raster("/Users/grace/rprojects/wrf_project/netcdfmaps_final/cellid.asc")
basin = raster("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")

# reproject DEM and cellid
dem_utm = projectRaster(dem, basin, method="bilinear")  # continuous data -> bilinear
writeRaster(dem_utm, "/Users/grace/rprojects/wrf_project/netcdfmaps_final/dem_utm.tif", overwrite=TRUE)

# load text ID 
id_df = read.table("/Users/grace/rprojects/wrf_project/netcdfmaps_final/xyid_loc.txt", header=FALSE)
colnames(id_df) = c("zone_id", "id_dup", "y", "x", "y_dup", "x_dup")
id_df = id_df[, c("x", "y", "zone_id")] # keep only what we need

# convert text file to spatial points -- this didn't work
# coordinates(id_df) = ~x+y
# proj4string(id_df) = crs("+proj=longlat +datum=WGS84")  # match DEM projection
# rasterize zone IDs, using basin as template to match extent, resolution, and CRS
# netcdf_msk = rasterize(id_df, dem, field="zone_id", fun="first")
# save raster
# writeRaster(netcdf_msk, "/Users/grace/rprojects/wrf_project/netcdfmaps_final/netcdf_msk.tif", format="GTiff", overwrite=TRUE)

# convert to terra vector, rasterize to netCDF grid
# id_vect = vect(id_df, geom=c("x","y"), crs="EPSG:4326")  # WGS84
# zone_raster = rasterize(id_vect, nc_grid_lonlat, field="zone_id", touches=TRUE)
# 
# # mask using basin
# basin = rast("/Users/grace/rprojects/wrf_project/SWbasin_AvgFire")
# basin_ll = project(basin, "EPSG:4326")  # project basin to lat/lon
# basin_resampled <- resample(basin_ll, zone_raster, method="near")
# zone_masked = mask(zone_raster, basin_resampled)
# 
# writeRaster(zone_masked, 
#             filename = "/Users/grace/rprojects/wrf_project/netcdfmaps_final/zone_masked.tif", 
#             filetype = "GTiff", overwrite = TRUE)

# check visually
plot(dem, main="DEM with NetCDF Zone Overlay and Basin Outline")
plot(netcdf_msk, add=TRUE, col=rainbow(230))

# create a data.frame linking patches to climate zone (optional) -- worldfile
# this apparently replaces the GRASS/basin2worldfile but the output didn't make sense
# patch_ids <- cellid
# 
# ## ensure values are aligned
# if (!all(dim(netcdf_msk) == dim(patch_ids))) {
#   patch_ids <- resample(patch_ids, netcdf_msk, method="ngb")
# }
# 
# worldfile <- data.frame(
#   patch_id = getValues(patch_ids),
#   zone_id  = getValues(netcdf_msk)
# )
# worldfile <- worldfile[!is.na(worldfile$patch_id), ]
# 
# # Save CSV
# write.csv(worldfile, "/Users/grace/rprojects/wrf_project/netcdfmaps_final/worldfile.csv", row.names=FALSE)

# Rasterize in WGS84 (same CRS as points)
netcdf_msk_wgs <- rasterize(id_df, dem, field="zone_id", fun="first")

# Reproject to UTM using nearest neighbor
netcdf_msk_utm_2 <- projectRaster(netcdf_msk_wgs, basin, method="ngb")
writeRaster(netcdf_msk_utm_2, "/Users/grace/rprojects/wrf_project/netcdfmaps_final/netcdf_msk_utm_2.tif", overwrite=TRUE)

id_vect <- vect(id_df) 
# Transform from WGS84 to NAD83 / UTM zone 10N
id_vect_utm <- project(id_vect, crs(basin))  # terra handles datum shift

# Optional: check bounding box
ext(id_vect_utm)
ext(basin)
```

This code below is from Will but I'm not sure what edits the Sedgwick WRF base station will need just yet -- none of this is in use it is just hanging out in orbit down below
```{r}
# ------------------------------ EDIT/FIX BASE STATION  ------------------------------

varnames = rep(NA, length(clim_files_trim))
for (i in seq_along(clim_files_trim)) {
  tmp = nc_open(clim_files_trim[i])
  varnames[i] = names(tmp$var)
}

ncbase = read.table(file.path(map_dest,"netcdf.base"))

ncbase[ncbase[,2]=="year_start_index", 1] = "1900"

ncbase[ncbase[,2]=="netcdf_tmax_filename", 1] = clim_files[grepl("tmmx", clim_files)]
ncbase[ncbase[,2]=="netcdf_var_tmax", 1] = varnames[grepl("tmmx", clim_files)]

ncbase[ncbase[,2]=="netcdf_tmin_filename", 1] = clim_files[grepl("tmmn", clim_files)]
ncbase[ncbase[,2]=="netcdf_var_tmin", 1] = varnames[grepl("tmmn", clim_files)]

ncbase[ncbase[,2]=="netcdf_rain_filename", 1] = clim_files[grepl("pr", clim_files)]
ncbase[ncbase[,2]=="netcdf_var_rain", 1] = varnames[grepl("pr", clim_files)]

ncbase[ncbase[,2]=="netcdf_huss_filename", 1] = clim_files[grepl("daily_mean_specific_humidity", varnames)]
ncbase[ncbase[,2]=="netcdf_var_huss", 1] = varnames[grepl("daily_mean_specific_humidity", varnames)]

ncbase[ncbase[,2]=="netcdf_rmax_filename", 1] = clim_files[grepl("rmax", clim_files)]
ncbase[ncbase[,2]=="netcdf_var_rmax", 1] = varnames[grepl("rmax", clim_files)]

ncbase[ncbase[,2]=="netcdf_rmin_filename", 1] = clim_files[grepl("rmin", clim_files)]
ncbase[ncbase[,2]=="netcdf_var_rmin", 1] = varnames[grepl("rmin", clim_files)]

ncbase[ncbase[,2]=="netcdf_rsds_filename", 1] = clim_files[grepl("daily_mean_shortwave_radiation_at_surface", varnames)]
ncbase[ncbase[,2]=="netcdf_var_rsds", 1] = varnames[grepl("daily_mean_shortwave_radiation_at_surface", varnames)]

ncbase[ncbase[,2]=="netcdf_was_filename", 1] = clim_files[grepl("daily_mean_wind_speed", varnames)]
ncbase[ncbase[,2]=="netcdf_var_was", 1] = varnames[grepl("daily_mean_wind_speed", varnames)]

write.table(ncbase, file.path(map_dest,"netcdf.base"), row.names = F, col.names = F, quote = F)
```

```{r}
library(terra)

# -------------------------------
# Inputs
# -------------------------------
nc_file  <- "/Users/grace/rprojects/wrf_project/dly_tmin/wrf.dlymin.airt2m.1994.nc"        
basin    <- rast("//Users/grace/rprojects/wrf_project/SWbasin_AvgFire") 

# -------------------------------
# Load NetCDF
# -------------------------------
r <- rast(nc_file)

# Sanity check
print(r)
# dimensions should be [rows, cols, time]

# -------------------------------
# Reduce time dimension
# -------------------------------
# OPTION A: mean over the year (most common)
r_year <- app(r, fun = mean, na.rm = TRUE)

r_proj <- project(r_year, basin, method = "bilinear")

# -------------------------------
# Clip to basin
# -------------------------------
r_clip <- mask(r_proj, basin)

# -------------------------------
# Save + visualize
# -------------------------------
#writeRaster(r_clip, out_tif, overwrite = TRUE)

plot(r_clip, main = "Average annual relative humidity in Sedgwick")
#plot(basin, add = TRUE, border = "black")
```

